Certified Symbolic Transducer with Applications in String Solving

### 2025-03-14 Paper/main.tex
finish second round revision

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{15th Conference on
Interactive Theorem Proving (ITP 2025)}
\EventShortTitle{ITP 2025}
\EventAcronym{ITP}
\EventYear{2025}
\EventDate{July 24--27, 2025}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite-state Automata (FAs) are fundamental components in the domains of programming languages. For instance, regular expressions, which are pivotal in languages such as JavaScript and Python, are frequently implemented using FAs.
%
Finite-state Transducers (FTs) extend the capabilities of FAs by enabling the transformation of input strings into output strings, thereby providing a more expressive framework for operations that encompass both recognition and transformation. Despite the widespread adoption of formalizing FAs and FTs within proof assistants such as Coq and Isabelle/HOL, these formalizations often fall short in terms of applicability to real-world scenarios. A significant limitation of classical finite-state models is that transition labels are typically confined to a single character from a finite alphabet. However, in practical applications, the alphabet of an FA or FT can be extensive or even \emph{infinite}. This classical approach to formalizing transitions can result in transition explosion, leading to critical performance bottlenecks of FA operations.

A more pragmatic approach involves the formalization of symbolic FAs \cite{cav/DAntoniV17} and FTs \cite{VeanesHLMB12Transducer}, where transition labels are symbolic and potentially infinite. While the formalization of symbolic FAs has been explored in the work of CertiStr \cite{cpp/KanLRS22}, the formalization of symbolic FTs in interactive proof assistants remains largely unexplored due to the increased complexity challenges.
%
In this paper, we aim to formalize symbolic FTs within the Isabelle/HOL framework. This formalization is refinement-based and is designed to be extensible with various symbolic representations of transition labels. To assess its performance, we applied the formalized symbolic FTs to an SMT string solver for modeling replacement operations. The experimental results indicate that the formalized symbolic transducer can efficiently solve string constraints.

@@

Even though there are various formalizations of FAs and FTs in interactive proof assistants such as Isabelle \cite{isabelle-homepage} and Coq \cite{coq-homepage}, these are predominantly based on classical definitions. However, these traditional approaches present certain limitations when applied to practical scenarios. One significant drawback is that transition labels are typically non-symbolic and finite. A conventional transition is represented as $q\xrightarrow{a}q'$, where $a$ is a character from a finite alphabet. This simplistic representation can lead to a phenomenon known as transition explosion. For example, if the alphabet $\Sigma$ encompasses the entire Unicode range, which is common in modern programming languages, it consists of $\texttt{0x110000}$ distinct characters. Defining a transition from state $q$ to $q'$ that accepts any character in $\Sigma$ would require splitting into $\texttt{0x110000}$ individual transitions, rendering the product of two FAs highly inefficient. Furthermore, in practical applications, it is often necessary to consider infinite alphabets, such as the set of all integers.

Symbolic Finite Automata (SFA) and Symbolic Finite-state Transducers (SFT) \cite{cav/DAntoniV17, VeanesHLMB12Transducer} represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic models utilize transition labels defined by boolean algebras, allowing for more expressive representations. For example, a transition label might be specified as an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from $'\texttt{a}'$ to $'\texttt{z}'$, or as an arithmetic condition ($x \% 2 = 0$), denoting all even numbers. This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

@@

Prior work has successfully formalized symbolic FAs within Isabelle/HOL through CertiStr \cite{cpp/KanLRS22}, demonstrating both the efficiency and effectiveness of symbolic FAs in practice. However, the formalization of FTs remains an open challenge, primarily due to their inherent complexity in two aspects: the formalization of transition labels and the specification of transition output functions. FTs constitute a significantly more expressive and powerful theoretical framework compared to symbolic FAs, as evidenced by their capability to model complex string transformations such as replacement operations.
%
Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.

In this work, we present a comprehensive formalization of symbolic FTs that builds upon the formalization of symbolic FAs. To address the extensibility challenges inherent in supporting diverse transition label theories, we adopt a refinement-based approach. At the abstraction level, transition labels are formalized through the fundamental mathematical concept of \emph{sets}. This abstraction facilitates subsequent refinement to various boolean algebras, including intervals and arithmetic conditions, while maintaining theoretical consistency.

@@

To evaluate the effectiveness of our formalization, we developed a string solver compliant with SMT-LIB language \cite{smtlib}, focusing particularly on replacement operations. We evaluated our implementation using a comprehensive set of benchmarks from the SMT-LIB repository \cite{smtlib_benchmarks}. The experimental results demonstrate that our formalization achieves computational efficiency in practical constraint-solving scenarios.

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite-state transducers in the Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodates various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop a comprehensive interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

@@

\begin{definition}[Symbolic Finite Transducer]
\label{def-sft}
   A Symbolic Finite Transducer over $\tau_1\rightarrow \tau_2$ is a quadruple $\mathcal{T} = (\mathcal{Q}, \Delta, \mathcal{I}, \mathcal{F})$, where 
   \begin{itemize}
   \item $\mathcal{Q}$ is a finite set of states,
   \item $\mathcal{I}\subseteq \mathcal{Q}$ is the set of initial states,
   \item $\mathcal{F} \subseteq\mathcal{Q}$ is the set of accepting states,
   \item $\Delta$ is the set of transition relations. Each element in $\Delta$ is of the form $(q, \phi, f, q')$ or written as $q\xrightarrow{\phi, f} q'$, where $q$ and $q'$ are states in $\mathcal{Q}$.
   $\phi$ is a predicate of type $\tau_1\rightarrow \mathbb{B}$.
   $f$ is a lambda term of type $\tau_1\rightarrow \tau_2$. $f$ is called an \emph{output function}.
   \end{itemize}

@@

SFTs accept an input word and generate an output. This can be defined by \emph{runs} of SFTs.
An SFT run $\sigma$ is a sequence $(q_0, \phi_0, f_0, q_1),(q_1, \phi_1, f_1, q_2),\ldots, (q_{n-1}, \phi_{n-1}, f_{n-1}, q_n)$ such that $q_0\in \mathcal{I}$ and $(q_i, 
\phi_i, f_i, q_{i+1}), 0 \leq i \leq n-1$ is a transition in $\Delta$.
$\sigma$ is an accepting run when $q_n$ is an accepting state.

@@

To make the operation \emph{product} more general, we extend the operation to an SFT and a set of input words represented by a regular language, which can be denoted by an NFA $\mathcal{A}$. More precisely,
\[
\mathcal{T}\times \mathcal{A} = \{w'\mid \exists w.~w\in \mathcal{L}(\mathcal{A})\land w' \in \mathcal{T}\times\{w\}\}, \text{ where }\mathcal{L}(\mathcal{A})\text{ denotes the language of }\mathcal{A}.
\]

@@

Firstly, $'a \text{ set option }$ is the input type of the transition, it accepts a set of elements of type $'a$ or $\texttt{None}$ corresponding to empty string $\varepsilon$. 
Accepting a set of $'a$ elements aims to express the same but more abstract semantics of the input labels in Definition \ref{def-sft}, in which an input label is a predicate. A predicate's semantics as introduced before represents a set of elements that make the predicate true. But predicates have various different forms. For instance, the interval $[1, 9]$ represents the set $\{e \mid 1 \leq e \leq 9\}$. The predicate $\lambda x \colon \mathbb{N}.~ x \text{ mod } 2 = 0$ denotes the set of even natural numbers. All these different forms of labels are abstracted as sets in our formalization.
%
The value $\texttt{None}$ represents the empty string $\varepsilon$ in our formalization, indicating a transition that consumes no input but may still produce output elements. This design choice facilitates the modeling of real-world applications in SFTs, as we will demonstrate in our application to string solvers.

@@

Figure \ref{fig-def-FTProd} depicts the abstract level formalization for the product of an SFT and an SFA. The parameters $\mathcal{T}$ and $\mathcal{A}$ are an SFT and an SFA, respectively. But we need to explain the role of $\text{\texttt{F}}$. The output function $f$ for each transition in $\Delta_t$ is of type "$'a\;\text{option} \Rightarrow 'b\;\text{set option}$", which applies to a \emph{single} element of type $'a$ or $\varepsilon$. $\text{\texttt{F}}$ extends $f$ to apply it to a set of elements. More precisely, let $f$ be an output function, the semantics of $\text{\texttt{F}}$ is defined as

@@

The transition relations $\Delta_e$ and $\Delta_e'$ are determinted by considering two distinct cases based on the nature of transitions in the SFT: $\varepsilon$-transitions and non-$\varepsilon$-transitions. In both cases, the transition labels in the resulting $\varepsilon$SFA are derived from the composition of the SFT's output function and the SFA's input labels:
For transitions in $\Delta_e$,

\begin{enumerate}
\item When $(p, (\text{None}, f), q)$ is a transition in $\Delta_t$, the input character is $\varepsilon$. Consequently, the SFA remains in its current state, and the product transition produces the output $(f~\text{None})$.

\item When $(p, (\text{Some}~\sigma_1, f), q)$ is a transition in $\Delta_t$, synchronization is possible only with SFA transitions that share characters with $\sigma_1$. The resulting output is $F~f~(\sigma_1\cap\sigma_2)$, where $\sigma_2$ represents the input label of the corresponding SFA transition.
\end{enumerate}

@@

The definition \texttt{outputL} generalizes \texttt{outputE} to characterize the set of all possible outputs that an SFT $\mathcal{T}$ can generate when processing inputs from the language accepted by the SFA $\mathcal{A}$. The reachability of a trace $\pi$ between states $q$ and $q'$ in an SFT $\mathcal{T}$ is verified by the predicate $\texttt{LTTS\_reachable}~\Delta_t~M~q~\pi~q'$.

@@

An interval is defined as a pair $(i, j)$ representing the set $\{e \mid i \leq e \leq j\}$. To achieve greater expressiveness, our formalization extends this notion to interval lists of the form $[(i_1, j_1), \ldots, (i_n, j_n)]$, which denote the set $\bigcup_{1\leq k\leq n}\{e \mid i_k \leq e \leq j_k\}$. This generalization offers two key advantages: it enables more compact representation of transitions in SFAs and SFTs through merging, and it allows for efficient handling of interval operations without unnecessary splitting. For example, the set difference between intervals $[1, 5]$ and $[3, 4]$ can be directly represented as the interval list $[(1, 2), (5, 5)]$.

@@

In this subsection, we present the algorithmic implementation of the product operation between an SFT and an SFA\footnote{For clarity of presentation, we show a simplified version of the Isabelle/HOL implementation while preserving the essential algorithmic structure.}. 
%
Figure \ref{fig-compute-nft-product} illustrates the core algorithm \texttt{productT\_impl}, which is implemented using the \emph{Refine\_Monadic} framework. Note that in the implementation, there are some new operations: \texttt{nfa\_states}, \texttt{nfa\_trans}, \texttt{nfa\_initial}, \texttt{nfa\_accepting}, and \texttt{nft\_tranfun}. There are corresponding to the states, transitions, initial states, accepting states, and output function mapping of the SFT.

@@

A central algorithmic challenge in our implementation lies in the computation of transition sets. As shown in Figure \ref{fig-def-prods-imp}, we implement this computation through the function \texttt{trans\_comp\_imp}, which computes the synchronization of transitions between the SFT and SFA. This function decomposes the synchronization process into two distinct cases, each handled by a specialized function:

@@

    We now present the implementation of \texttt{subtrans\_comp} in detail, as shown in Figure \ref{fig-def-subtrans_comp} (the implementation of \texttt{subtrans\_comp\_$\varepsilon$} follows analogous principles). This function systematically processes transitions in the SFA, represented by the set \texttt{T}. For each transition $(q_1, \alpha', q_1')\in \texttt{T}$, the function performs two key operations when the intersection of input labels is non-empty ($\alpha \cap \alpha' \neq \emptyset$, verified using \texttt{nemptyIs}):

    \begin{enumerate}
      \item Computes non-$\varepsilon$-transitions ($\texttt{D1}$): When the output function applied to the intersection $\alpha \cap \alpha'$ yields a non-empty set, a new transition is added to $\texttt{D1}$ with the computed output label.
      \item Generates $\varepsilon$-transitions ($\texttt{D2}$): When there exists an input in the intersection $\alpha \cap \alpha'$ that produces an empty string (verified by checking if $\texttt{M}~f$ maps any element to \texttt{None}. It is implemented by \texttt{fe}), a corresponding $\varepsilon$-transition is added to $\texttt{D2}$.
    \end{enumerate}

@@

The string replacement operation, denoted as $\texttt{replace}(\texttt{str}, \texttt{pattern}, \texttt{replacement})$, is a fundamental string transformation that takes three parameters:
\begin{itemize}
  \item $\texttt{str}$: The input string to be transformed
  \item $\texttt{pattern}$: A regular expression defining the matching criteria
  \item $\texttt{replacement}$: The string to be substituted for matched substrings
\end{itemize}

@@

\begin{enumerate}
  \item First, construct an SFA that recognizes the regular expression pattern $\texttt{/[0-9]+/}$. Transform this SFA into an SFT by augmenting each transition with an output function $f = \lambda x.~\texttt{None}$ that produces the empty string. 
  \item Second, construct an SFA that accepts the replacement string "\texttt{NUM}". Since a constant string can be viewed as a specialized regular expression, we can construct its SFA representation. Convert this SNFA into an SFT by adding $\varepsilon$-transitions  and appropriate output functions that emit the characters of "\texttt{NUM}" in sequence.
  \item Finally, compose the two SFTs through concatenation and augment the resulting transducer with self-loop transitions at the initial and final states. These additional transitions, labeled with $\Sigma$ (the set of all characters in the alphabet) and empty output function, enable the SFT to process arbitrary prefixes and suffixes of the input string while preserving the replacement behavior on matched substrings.
\end{enumerate}

@@

   \path[->] 
   (q0') edge [bend left] node {\texttt{[0-9]}/$\texttt{f}$} (q1')
   (q1') edge [loop above] node {\texttt{[0-9]}/$\texttt{f}$} (q1');
\end{tikzpicture}
\caption{Corresponding SFA and SFT for \texttt{/[0-9]+/}}
\label{fig-snfa-pattern}
\end{figure}

@@

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=1.8cm, auto]
  % Original NFA for "ITP"
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state] (q1) [right=of q0] {$q_1$}; 
  \node[state] (q2) [below=of q0] {$q_2$}; 
  \node[state, accepting] (q3) [right=of q2] {$q_3$};

@@

  \path[->] 
  (q0') edge node {\texttt{None}/\texttt{g 1}} (q1')
  (q1') edge node {\texttt{None}/\texttt{g 2}} (q2')
  (q2') edge node[below] {\texttt{None}/\texttt{g 3}} (q3');
\end{tikzpicture}
\caption{Corresponding SFA and SFT for "\texttt{NUM}"}
\label{fig-snfa-replacement}
\end{figure}

\noindent\emph{Step 3.}
Figure \ref{fig-rearranged-automata} presents the complete SFT construction, obtained by composing the pattern-matching SFT (Fig. \ref{fig-snfa-pattern}) with the replacement-generating SFT (Fig. \ref{fig-snfa-replacement}). The composition process involves two key modifications:
\begin{enumerate}
  \item Connect the two SFTs by adding $\varepsilon$-transitions (labeled with \texttt{None}/\texttt{f}) from each accepting state of the pattern-matching SFT to the initial state of the replacement-generating SFT
  \item Augment the resulting transducer with self-loop transitions at both ends, labeled with $\Sigma/\texttt{f}$, where $\Sigma$ represents the full alphabet. These transitions enable the SFT to process arbitrary input prefixes and suffixes while preserving the matched substring for replacement
\end{enumerate}

\begin{figure}[h] \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=1.8cm, auto]
    % Second automaton from the first figure
    \node[state, initial] (q0') {$q_0'$}; 
    \node[state] (q1') [right=of q0'] {$q_1'$};

     \path[->] 
     (q0') edge [bend left] node {\texttt{[0-9]}/$\texttt{f}$} (q1')
     (q1') edge [loop above] node {\texttt{[0-9]}/$\texttt{f}$} (q1')
     (q0') edge [loop below] node {$\Sigma$/$\texttt{f}$} (q0');

>>>>

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{16th Conference on
Interactive Theorem Proving (ITP 2025)}
\EventShortTitle{ITP 2025}
\EventAcronym{ITP}
\EventYear{2025}
\EventDate{July 24--27, 2025}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite-state Automata (FAs) are fundamental components in the domains of programming languages. For instance, regular expressions, which are pivotal in languages such as JavaScript and Python, are frequently implemented using FAs.
%
Finite-state Transducers (FTs) extend the capabilities of FAs by enabling the transformation of input strings into output strings, thereby providing a more expressive framework for operations that encompass both recognition and transformation. Despite the widespread adoption of formalizing FAs and FTs within proof assistants such as Coq and Isabelle/HOL, these formalizations often fall short in terms of applicability to real-world scenarios. A significant limitation of classical finite-state models is that transition labels are typically defined as a single character from a finite alphabet. However, in practical applications, the alphabet of an FA or FT can be extensive or even \emph{infinite}. This classical approach to formalizing transitions can result in transition explosion, leading to critical performance bottlenecks of FA and FT operations.

A more pragmatic approach involves the formalization of symbolic FAs \cite{cav/DAntoniV17} and FTs \cite{VeanesHLMB12Transducer}, where transition labels are symbolic and potentially infinite. While the formalization of symbolic FAs has been explored in the work of CertiStr \cite{cpp/KanLRS22}, the formalization of symbolic FTs in interactive proof assistants remains largely unexplored due to the increased complexity challenges.
%
In this paper, we aim to formalize symbolic FTs within the Isabelle/HOL framework. This formalization is refinement-based and is designed to be extensible with various symbolic representations of transition labels. To assess its performance, we applied the formalized symbolic FTs to an SMT string solver for modeling replacement operations. The experimental results indicate that the formalized symbolic transducer can efficiently and effectively solve string constraints with replacement operations.

@@

Even though there are various formalizations of FAs and FTs in interactive proof assistants such as Isabelle \cite{isabelle-homepage} and Coq \cite{coq-homepage}, these are predominantly based on classical definitions. However, these traditional approaches present certain limitations when applied to practical scenarios. One significant drawback is that transition labels are typically non-symbolic and finite. A conventional transition is represented as $q\xrightarrow{a}q'$, where $a$ is a character from a finite alphabet. This simplistic representation can lead to a phenomenon known as transition explosion. For example, if the alphabet $\Sigma$ encompasses the entire Unicode range, which is common in modern programming languages, it consists of $\texttt{0x110000}$ distinct characters. Defining a transition from state $q$ to $q'$ that accepts any character in $\Sigma$ would require splitting into $\texttt{0x110000}$ individual transitions, rendering the FA and FT operations highly inefficient. Furthermore, in practical applications, it is often necessary to consider infinite alphabets, such as the set of all integers.

Symbolic Finite-state Automata (SFA) and Symbolic Finite-state Transducers (SFT) \cite{cav/DAntoniV17, VeanesHLMB12Transducer} represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic models utilize transition labels defined by boolean algebras, allowing for more expressive representations. For example, a transition label might be specified as an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from $'\texttt{a}'$ to $'\texttt{z}'$, or as an arithmetic condition ($x \mod 2 = 0$), denoting all even numbers. This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

@@

Prior work of CertiStr \cite{cpp/KanLRS22} has successfully formalized SFAs within Isabelle/HOL, demonstrating both the efficiency and effectiveness of SFAs in practice. However, the formalization of FTs remains an open challenge, primarily due to their inherent complexity in two aspects: the formalization of transition labels and the specification of transition output functions. FTs constitute a significantly more expressive and powerful theoretical framework compared to symbolic FAs, as evidenced by their capability to model complex string transformations such as replacement operations.
%
Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.

In this work, we present a comprehensive formalization of SFTs. To address the extensibility challenges inherent in supporting diverse transition label theories, we adopt a refinement-based approach. At the abstraction level, transition labels are formalized through the fundamental mathematical concept of \emph{sets}. This abstraction facilitates subsequent refinement to various boolean algebras, including intervals and arithmetic conditions, while maintaining theoretical consistency.

@@

To evaluate the effectiveness and efficiency of our formalization, we developed a string solver compliant with SMT-LIB language \cite{smtlib}, focusing particularly on replacement operations. We evaluated our implementation using a set of benchmarks from the SMT-LIB repository \cite{smtlib_benchmarks}. The experimental results demonstrate that our formalization achieves computational efficiency in constraint-solving scenarios.

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite-state transducers in the Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodates various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop an interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

@@

\begin{definition}[Symbolic Finite-state Transducer]
\label{def-sft}
   A Symbolic Finite-state Transducer over $\tau_1\rightarrow \tau_2$ is a quadruple $\mathcal{T} = (\mathcal{Q}, \Delta, \mathcal{I}, \mathcal{F})$, where 
   \begin{itemize}
   \item $\mathcal{Q}$ is a finite set of states,
   \item $\mathcal{I}\subseteq \mathcal{Q}$ is the set of initial states,
   \item $\mathcal{F} \subseteq\mathcal{Q}$ is the set of accepting states,
   \item $\Delta$ is the set of transition relations. Each element in $\Delta$ is of the form $(q, \phi, f, q')$ or written as $q\xrightarrow{\phi, f} q'$, where $q$ and $q'$ are states in $\mathcal{Q}$.
   $\phi$ is a predicate of type $\tau_1\rightarrow \mathbb{B}$.
   $f$ is a lambda term of type $\tau_1\rightarrow \tau_2$. $f$ is called an \emph{output function}.
   \end{itemize}

@@

SFTs accept an input word and generate an output word. This can be defined by \emph{runs} of SFTs.
An SFT run $\sigma$ is a sequence $(q_0, \phi_0, f_0, q_1),(q_1, \phi_1, f_1, q_2),\ldots, (q_{n-1}, \phi_{n-1}, f_{n-1}, q_n)$ such that $q_0\in \mathcal{I}$ and $(q_i, 
\phi_i, f_i, q_{i+1}), 0 \leq i \leq n-1$ is a transition in $\Delta$.
$\sigma$ is an accepting run when $q_n$ is an accepting state.

@@

To make the operation \emph{product} more general, we extend the operation to an SFT and a set of input words represented by a regular language, which can be denoted by an SFA $\mathcal{A}$. More precisely,
\[
\mathcal{T}\times \mathcal{A} = \{w'\mid \exists w.~w\in \mathcal{L}(\mathcal{A})\land w' \in \mathcal{T}\times\{w\}\}, \text{ where }\mathcal{L}(\mathcal{A})\text{ denotes the language of }\mathcal{A}.
\]

@@

Firstly, $'a \text{ set option }$ is the input type of the transition, it accepts a set of elements of type $'a$ or $\texttt{None}$ corresponding to empty string $\varepsilon$. 
Accepting a set of $'a$ elements aims to express the same but more abstract semantics of the input labels in Definition \ref{def-sft}, in which an input label is a predicate. A predicate's semantics as introduced before represents a set of elements that make the predicate true. But predicates have various different forms. For instance, the interval $[1, 9]$ represents the set $\{e \mid 1 \leq e \leq 9\}$. The predicate $\lambda b.~ b[7] = 1$, where $b$ is a bit vector of length 8, denotes the set of bit vectors that have a 1 in the 7th position. All these different forms of labels are abstracted as sets in our formalization.
%
The value $\texttt{None}$ represents the empty string $\varepsilon$ in our formalization, indicating a transition that consumes no input but may still produce output elements. This design choice facilitates the modeling of real-world applications in SFTs, as we will demonstrate in our application to string solvers.

@@

Figure \ref{fig-def-FTProd} depicts the abstract level formalization for the product of an SFT and an SFA. The parameters $\mathcal{T}$ and $\mathcal{A}$ are an SFT and an SFA, respectively. The 
result of the product operation is an $\varepsilon$SFA.
But we need to explain the role of $\text{\texttt{F}}$. The output function $f$ for each transition in $\Delta_t$ is of type "$'a\;\text{option} \Rightarrow 'b\;\text{set option}$", which applies to a \emph{single} element of type $'a$ or $\varepsilon$. $\text{\texttt{F}}$ extends $f$ to apply it to a set of elements. More precisely, let $f$ be an output function, the semantics of $\text{\texttt{F}}$ is defined as

@@

The transition relations $\Delta_e$ and $\Delta_e'$ are determinted by considering two distinct cases based on the nature of transitions in the SFT: $\varepsilon$-transitions and non-$\varepsilon$-transitions. In both cases, the transition labels in the resulting $\varepsilon$SFA are derived from the composition of the SFT's output function and the SFA's input labels:

\begin{enumerate}
\item When $(p, (\text{None}, f), q)$ is a transition in $\Delta_t$, i.e. the input character is $\varepsilon$. Consequently, the SFA $\mathcal{A}$ remains in its current state, and the product transition produces the output $(f~\text{None})$.

\item When $(p, (\text{Some}~\sigma_1, f), q)$ is a transition in $\Delta_t$, synchronization is possible only with SFA transitions that share characters with $\sigma_1$, i.e., $\sigma_1\cap\sigma_2\neq \emptyset$. The resulting output is $F~f~(\sigma_1\cap\sigma_2)$, where $\sigma_2$ represents the input label of the corresponding SFA transition.
\end{enumerate}

@@

The definition \texttt{outputL} generalizes \texttt{outputE} to characterize the set of all possible outputs that an SFT $\mathcal{T}$ can generate when processing inputs from the language accepted by the SFA $\mathcal{A}$. The reachability of a trace $\pi$ between states $q$ and $q'$ in an SFT $\mathcal{T}$ is verified by the predicate $\texttt{LTTS\_reachable}~\Delta_t~\texttt{M}~q~\pi~q'$.

@@

An interval is defined as a pair $(i, j)$ representing the set $\{e \mid i \leq e \leq j\}$. To achieve greater expressiveness, our formalization extends this notion to interval lists of the form $[(i_1, j_1), \ldots, (i_n, j_n)]$, which denote the set $\bigcup_{1\leq k\leq n}\{e \mid i_k \leq e \leq j_k\}$. This generalization offers two key advantages: it enables more compact representation of transitions in SFAs and SFTs through merging, and it allows for efficient handling of interval operations without unnecessary splitting. For example, the set difference between intervals $(1, 5)$ and $(3, 4)$ can be directly represented as the interval list $[(1, 2), (5, 5)]$.

@@

In this subsection, we present the algorithmic implementation of the product operation between an SFT and an SFA\footnote{For clarity of presentation, we show a simplified version of the Isabelle/HOL implementation while preserving the essential algorithmic structure.}. 
%
Figure \ref{fig-compute-nft-product} illustrates the core algorithm \texttt{productT\_impl}, which is implemented using the \emph{Refine\_Monadic} framework. Note that in the implementation, there are some refined operations: \texttt{nfa\_states}, \texttt{nfa\_trans}, \texttt{nfa\_initial}, \texttt{nfa\_accepting}, and \texttt {nft\_tranfun}. These are corresponding to the states, transitions, initial states, accepting states, and output function mapping of the SFT.

@@

A central algorithmic challenge in our implementation lies in the computation of transition sets $\texttt{D1}$ and $\texttt{D2}$ in Figure \ref{fig-compute-nft-product}. As shown in Figure \ref{fig-def-prods-imp}, we implement this computation through the function \texttt{trans\_comp\_imp}, which computes the synchronization of transitions between the SFT and SFA. This function decomposes the synchronization process into two distinct cases, each handled by a specialized function:

@@

    We now present the implementation of \texttt{subtrans\_comp} in detail, as shown in Figure \ref{fig-def-subtrans_comp} (the implementation of \texttt{subtrans\_comp\_$\varepsilon$} follows analogous principles). For a transtion $(q, \alpha, f, q')$ in the SFT, this function traverses all transitions in the SFA, represented by the set \texttt{T}. For each transition $(q_1, \alpha', q_1')\in \texttt{T}$, the function performs two key operations when the intersection of input labels is non-empty ($\alpha \cap \alpha' \neq \emptyset$, verified using \texttt{nemptyIs}):

    \begin{enumerate}
      \item Computes non-$\varepsilon$-transitions ($\texttt{D1}$): When the output function applied to the intersection $\alpha \cap \alpha'$ yields a non-empty set, a new transition is added to $\texttt{D1}$ with the computed output label.
      \item Generates $\varepsilon$-transitions ($\texttt{D2}$): When there exists at least one input in the intersection $\alpha \cap \alpha'$ that produces an empty string (verified by checking if $\texttt{M}~f$ maps any element to \texttt{None}. It is implemented by \texttt{fe}), a corresponding $\varepsilon$-transition is added to $\texttt{D2}$.
    \end{enumerate}

@@

The string replacement operation, denoted as $\texttt{replace}(\texttt{str}, \texttt{pattern}, \texttt{replacement})$, is a fundamental string transformation that takes three parameters:
\begin{itemize}
  \item $\texttt{str}$: The input string to be transformed.
  \item $\texttt{pattern}$: A regular expression defining the matching criteria.  
  \item $\texttt{replacement}$: The string to be substituted for the matched substring.
\end{itemize}

@@

\begin{enumerate}
  \item First, construct an SFA that recognizes the regular expression pattern $\texttt{/[0-9]+/}$. Transform this SFA into an SFT by augmenting each transition with an output function $f = \lambda x.~\texttt{None}$ that produces the empty string. 
  \item Second, construct an SFA that accepts the replacement string "\texttt{NUM}". Since a constant string can be viewed as a specialized regular expression, we can construct its SFA representation. Convert this SFA into an SFT by adding $\varepsilon$-transitions  and appropriate output functions that emit the characters of "\texttt{NUM}" in sequence.
  \item Finally, compose the two SFTs through concatenation and augment the resulting transducer with self-loop transitions at the initial and final states. These additional transitions, labeled with $\Sigma$ (the set of all characters in the alphabet) and empty output function, enable the SFT to process arbitrary prefixes and suffixes of the input string while preserving the replacement behavior on matched substrings.
\end{enumerate}

@@

   \path[->] 
   (q0') edge [bend left] node {\texttt{[0-9]},$\texttt{f}$} (q1')
   (q1') edge [loop above] node {\texttt{[0-9]},$\texttt{f}$} (q1');
\end{tikzpicture}
\caption{Corresponding SFA and SFT for \texttt{/[0-9]+/}}
\label{fig-snfa-pattern}
\end{figure}

@@

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
  % Original NFA for "ITP"
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state] (q1) [right=of q0] {$q_1$}; 
  \node[state] (q2) [below=of q0] {$q_2$}; 
  \node[state, accepting] (q3) [right=of q2] {$q_3$};

@@

  \path[->] 
  (q0') edge node {\texttt{None}, \texttt{(g 1)}} (q1')
  (q1') edge node {\texttt{None}, \texttt{(g 2)}} (q2')
  (q2') edge node[below] {\texttt{None}, \texttt{(g 3)}} (q3');
\end{tikzpicture}
\caption{Corresponding SFA and SFT for "\texttt{NUM}"}
\label{fig-snfa-replacement}
\end{figure}

\noindent\emph{Step 3.}
Figure \ref{fig-rearranged-automata} presents the complete SFT construction, obtained by composing the pattern-matching SFT (Fig. \ref{fig-snfa-pattern}) with the replacement-generating SFT (Fig. \ref{fig-snfa-replacement}). The composition process involves two key modifications:
\begin{enumerate}
  \item Connect the two SFTs by adding $\varepsilon$-transitions (labeled with "\texttt{None}, \texttt{f}") from each accepting state of the pattern-matching SFT to the initial state of the replacement-generating SFT
  \item Augment the resulting transducer with self-loop transitions at both ends, labeled with "$\Sigma,~\texttt{f}$", where $\Sigma$ represents the full alphabet. For the string solver, $\Sigma$ is the set of all unicode characters. These transitions enable the SFT to process arbitrary input prefixes and suffixes while preserving the matched substring for replacement
\end{enumerate}

\begin{figure}[h] \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
    % Second automaton from the first figure
    \node[state, initial] (q0') {$q_0'$}; 
    \node[state] (q1') [right=of q0'] {$q_1'$};

     \path[->] 
     (q0') edge [bend left] node {\texttt{[0-9]}, $\texttt{f}$} (q1')
     (q1') edge [loop above] node {\texttt{[0-9]}, $\texttt{f}$} (q1')
     (q0') edge [loop below] node {$\Sigma$, $\texttt{f}$} (q0');
